---
title: "STAT 405 - Project Report"
author: "Isobel Adams & Danial Tajabadipour"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

The Vancouver Police Department has been counting crime counts for specific neighborhoods, the publically availble crime data is presented in an pdf found at: 

`https://vpd.ca/wp-content/uploads/2026/01/neighbourhood-stats-year-end-2025.pdf`

The dataset contains reported counts of several crime categories across Vancouver neighbourhoods, including sex offences, assaults, robbery, break and enter, theft of motor vehicle, theft from auto, theft under $5000, arson, mischief, and offensive weapons.

Because the data are published in PDF format rather than as a directly downloadable CSV, we manually entered the table into an R dataframe.

In our analysis we hope to use Bayesian analysis to quantify crime rates for different types of crimes, and whether we can predict future crime counts for selected neighbourhoods and crime types. 

The motivation behind our study is to hopefully find ways in which crime can be prevented, for instance if we find that crime rates are higher in a specific neighbourhood then crime rates could be decreased by enforcing stricter policing in that area.

```{r}
crime_data <- data.frame(
  neighbourhood = c(
    "Arbutus Ridge","Central Business District","Dunbar-Southlands","Fairview",
    "Grandview-Woodland","Hastings-Sunrise","Kensington-Cedar Cottage",
    "Kerrisdale","Killarney","Kitsilano","Marpole","Mount Pleasant","Musqueam",
    "Oakridge","Renfrew-Collingwood","Riley Park","Shaughnessy","South Cambie",
    "Stanley Park","Strathcona","Sunset","Victoria-Fraserview","West End",
    "West Point Grey"),
  sex_offences = c( 0,151,8,16,27,29,34,6,14,12,8,24,2,8,30,12,0,13,1,43,24,15,43,4),
  assaults = c(18,1591,22,115,254,97,185,22,51,101,69,215,15,26,168,41,15,24,18,622,80,
               45,306,24),
  robbery = c( 1,218,0,9,22,7,26,3,3,10,4,36,0,0,22,3,2,2,4,64,9,9,35,2),
  break_enter = c(25,438,24,164,121,84,127,29,37,92,89,173,0,44,123,75,32,33,12,
                  151,77,29,149,29),
  theft_mv = c(7,74,7,16,64,31,48,10,24,18,22,28,0,7,53,13,6,6,2,48,36,24,23,5),
  theft_auto = c(41,1855,72,219,219,222,317,73,84,202,146,220,0,56,202,148,43,
                 56,54,369,148,136,443,60),
  theft_under_5k = c(69,4458,70,706,507,312,601,132,159,465,276,1015,
                     1,115,1446,178,38,140,53,603,637,125,1120,77),
  arson = c(1,68,1,6,15,9,8,0,1,9,1,20,0,0,8,5,0,2,2,37,4,3,10,0),
  mischief = c(34,1864,33,209,281,140,209,45,83,142,134,296,6,29,171,86,28,42,16,434,137,
               85,318,32),
  offensive_weapons = c(4,708,4,28,44,9,25,5,5,10,8,39,0,1,9,2,3,4,4,244,9,10,38,3))
```

```{r}
head(crime_data)
```

## Models

Since the data are counts, we model $y_{ij}$ using Poisson distributions to estimate underlying crime intensity.

We consider two models.

**Model 1:** Simpler model

We treat each neighbourhood-crime combination independently:

$$
y_{ij} \mid \lambda_{ij} \sim \text{Poisson}(\lambda_{ij}), \quad
\lambda_{ij} \sim \text{Gamma}(\alpha, \beta)
$$

The Gamma prior is conjugate to the Poisson likelihood, allowing straightforward posterior computation.

**Model 2:** Hierarchical model

We incorporate neighbourhood and crime-type effects:

$$
y_{ij} \mid \lambda_{ij} \sim \text{Poisson}(\lambda_{ij}), \quad
\log(\lambda_{ij}) = \mu + \alpha_i + \beta_j
$$

$$
\alpha_i \sim \mathcal{N}(0, \sigma_\alpha^2), \quad
\beta_j \sim \mathcal{N}(0, \sigma_\beta^2)
$$

where $\mu$ is an overall intercept, $\alpha_i$ is the neighbourhood effect, and $\beta_j$ is the crime-type effect.

We use Gaussian priors for $\alpha_i$ and $\beta_j$, reflecting that most effects are near zero unless supported by the data. The variance parameters $\sigma_\alpha^2$ and $\sigma_\beta^2$ control variability across groups.

To ensure identifiability, we impose
$$
\sum_i \alpha_i = 0, \qquad \sum_j \beta_j = 0.
$$

We interpret $\lambda_{ij}$ as the expected number of crimes per year for crime type $j$ in neighbourhood $i$, representing a yearly rate without explicit adjustment for exposure or opportunity.

