---
title: "Stat 405 -Project Proposal"
author: "Isobel Adams & Danial Tajabadipour"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Public Github Repo: https://github.com/isobela/Stat-405-Project-Proposal

To ensure both team members contribute roughly equally, we plan to divide the work into data preparation / exploratory analysis and Bayesian modeling / posterior computation, with both members contributing to the final report writing, interpretation, and code review.

## Data

The Vancouver Police Department has been counting crime counts for specific neighborhoods, the publically availble crime data is presented in an pdf found at: 

`https://vpd.ca/wp-content/uploads/2026/01/neighbourhood-stats-year-end-2025.pdf`

The dataset contains reported counts of several crime categories across Vancouver neighbourhoods, including sex offences, assaults, robbery, break and enter, theft of motor vehicle, theft from auto, theft under $5000, arson, mischief, and offensive weapons.

Because the data are published in PDF format rather than as a directly downloadable CSV, we manually entered the table into an R dataframe.

As a backup dataset, we plan to use another year of the Vancouver Police Department neighbourhood crime statistics, so that the theme and data structure remain comparable.

Backup dataset: https://vpd.ca/wp-content/uploads/2025/01/neighbourhood-year-end-2024.pdf

```{r}
crime_data <- data.frame(
  neighbourhood = c(
    "Arbutus Ridge","Central Business District","Dunbar-Southlands","Fairview",
    "Grandview-Woodland","Hastings-Sunrise","Kensington-Cedar Cottage",
    "Kerrisdale","Killarney","Kitsilano","Marpole","Mount Pleasant","Musqueam",
    "Oakridge","Renfrew-Collingwood","Riley Park","Shaughnessy","South Cambie",
    "Stanley Park","Strathcona","Sunset","Victoria-Fraserview","West End",
    "West Point Grey"),
  sex_offences = c( 0,151,8,16,27,29,34,6,14,12,8,24,2,8,30,12,0,13,1,43,24,15,43,4),
  assaults = c(18,1591,22,115,254,97,185,22,51,101,69,215,15,26,168,41,15,24,18,622,80,
               45,306,24),
  robbery = c( 1,218,0,9,22,7,26,3,3,10,4,36,0,0,22,3,2,2,4,64,9,9,35,2),
  break_enter = c(25,438,24,164,121,84,127,29,37,92,89,173,0,44,123,75,32,33,12,
                  151,77,29,149,29),
  theft_mv = c(7,74,7,16,64,31,48,10,24,18,22,28,0,7,53,13,6,6,2,48,36,24,23,5),
  theft_auto = c(41,1855,72,219,219,222,317,73,84,202,146,220,0,56,202,148,43,
                 56,54,369,148,136,443,60),
  theft_under_5k = c(69,4458,70,706,507,312,601,132,159,465,276,1015,
                     1,115,1446,178,38,140,53,603,637,125,1120,77),
  arson = c(1,68,1,6,15,9,8,0,1,9,1,20,0,0,8,5,0,2,2,37,4,3,10,0),
  mischief = c(34,1864,33,209,281,140,209,45,83,142,134,296,6,29,171,86,28,42,16,434,137,
               85,318,32),
  offensive_weapons = c(4,708,4,28,44,9,25,5,5,10,8,39,0,1,9,2,3,4,4,244,9,10,38,3))
```

```{r}
head(crime_data)
```


## Bayesian Questions
Using the `crime_data` we hope to answer the following Bayesian questions:
 
* What are the underlying rates of different categories of crime across Vancouver neighbourhoods?
* What future crime counts might be predicted for selected neighbourhood and crime-type combinations?

## Methodolgy / Plan

Since we are working with count data we will assume crime counts $y_{ij}$ to be i.i.d for each crime type and neighbourhood. We hope to find true crime rate for each type of crime and neighborhood.

We will use two models

**Model 1**: Simpler model

As a first model we will treat each neighbourhood-crime combination independently:

$$
y_{ij} | \lambda_{ij} \sim \text{Poisson}(\lambda_{ij})\\
\lambda_{ij} \sim \text{Gamma}(\alpha, \beta)
$$
where $y_{ij}$ represents the crime count in neighborhood $i$ for crime type $j$. The Gamma prior is conjugate to the Poisson likelihood hence it is an reasonable choice for prior. However we may choose to change this when we continue with exploratory data analysis.

**Model 2**: Hierarchical Model

As a less naive model we will incorporate neighbourhood and crime-type effects, this will also make the model less sensitive to prior choice. $\lambda_{ij}$ must be positive hence a possible model could be

$$
y_{ij} | \lambda_{ij} \sim \text{Poisson}(\lambda_{ij})\\
\log(\lambda_{ij}) = \mu + \alpha_i + \beta_j \\
\alpha \sim  \space ?? \\
\beta \sim  \space ??
$$
where $\mu$ is an overall intercept, $\alpha$ is effect of neighborhood $i$ and $\beta$ is effect of crime type $j$. We will place reasonable priors on $\alpha$ and $\beta$.

The project requires two different techniques of posterior computation, we plan to use the following computational methods:

1. SNIS
2. MCMC

Using techniques from course to compare these two approaches.


